A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-03 text/discussion.Rmd
double check results, updated text in results and discussion section. added ubiq-proteasome gene annotation on right side. added figure s5

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated genesets related to diseases, syndromes, or classes
of proteins ( _e.g._ transcription factors) which are potentially problematic
[@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
already high in the remaining pathways for some driver genes, and to exclude
biologically unrelated collections of genes. We do not account for non-coding
mutations, amplification events or structural variants, though we hope to
include them in subsequent studies (see below). Our pathway-centric analysis
also allows us to circumvent the problem that some cancer-promoting mutations
likely occur at low enough frequencies as to be indistinguishable from
background. Previous attempts to capture these associations better using
Bayesian networks have yielded important new insights into this problem and new
genes [@hornNetSigNetworkbasedDiscovery2018] that would not have been possible
with traditional analysis, supporting the case that pathway-centric approaches
like ours have the potential to increase statistical power.

@@

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but are not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

>>>>

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain, untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated genesets related to diseases, syndromes, or classes
of proteins ( _e.g._ transcription factors) which are potentially problematic
[@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
already high in the remaining pathways for some driver genes, and to exclude
biologically unrelated collections of genes. We do not account for non-coding
mutations, amplification events or structural variants, though we hope to
include them in subsequent studies (see below). Our pathway-centric analysis
also allows us to circumvent the problem that some cancer-promoting mutations
likely occur at low enough frequencies as to be indistinguishable from
background. Previous attempts to capture these associations better using
Bayesian networks have yielded important new insights into this problem and new
genes [@hornNetSigNetworkbasedDiscovery2018] that would not have been possible
with traditional analysis, supporting the case that pathway-centric approaches
like ours have the potential to increase statistical power.

@@

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant clusters 8, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but are not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 10
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

### 2020-08-03 text/manuscript.Rmd
double check results, updated text in results and discussion section. added ubiq-proteasome gene annotation on right side. added figure s5

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
library(cowplot)
library(doRNG)
library(uwot)
library(janitor)
library(FactoMineR)
library(seriation)
library(rtracklayer)
library(poibin)
```

@@

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to start with somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
region, and splicing mutations. In order to minimize bias from well-studied diseases
and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
of interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total
of 9040 genes; see methods for details).

@@

Since pathways share some common driver genes, we investigated whether
tumors in each class  have multiple mutations across signaling pathways, or
whether a small number of driver genes account for enrichment. The split between
clusters 7 and 8 is largely driven by mutations in _PI3K_ and its
homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
**supplementary figure S4**). Furthermore, cluster 1 is distinguished
from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
_TP53_ panel of **Supplementary Figure S4**).

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways (**figure 3**) we applied this method to metastatic vs non-metastatic
tumors, as we find that metastatic tumors are distributed across all 12 clusters
( _e.g._ **figure 1l**). Using all non-metastatic tumors as background, we found
very low levels of enrichment, less than $10\%$, in a handful of pathways. We
reasoned that the individual clusters might be too different to detect global
metastasis enrichment signal given the small sample size (n = 215 metastatic
tumor samples).

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found 99 enriched pathways across all clusters (**table 2**). A number of
enrichments were found in multiple clusters represented by pathways that were
already shown to be enriched in a neighboring cluster. For example,
“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
Cluster 1, one of two clusters for which we could not identify any
distinguishing pathways (cluster 8 being the other), had enrichment in two
pathways that were otherwise specific to non-metastatic tumors of other
clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
samples were enriched for two key neutrophil pathways, “Neutrophil
Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.

```{r child='../code/table2.Rmd'}
```

There were also a number of pathways that were enriched with lower specificity
(among 4 or more clusters) which were found to be enriched in metastases. We
identified 18 such pathways of which 12 had significantly higher mutation rates
in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
dependent signaling in response to WNT”, “Class I MHC mediated antigen
processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**).

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin)
))
```

@@

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to start with somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
region, and splicing mutations. In order to minimize bias from well-studied diseases
and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
of interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total
of 8940 genes; see methods for details).

@@

Since pathways share some common driver genes, we investigated whether
tumors in each class  have multiple mutations across signaling pathways, or
whether a small number of driver genes account for enrichment. The split between
clusters 7 and 8 is largely driven by mutations in _PI3K_ and its
homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
**supplementary figure S5**). Furthermore, cluster 1 is distinguished
from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
_TP53_ panel of **Supplementary Figure S5**).

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways (**figure 2**) we applied this method to metastatic vs non-metastatic
tumors, as we find that metastatic tumors are distributed across all ten clusters
( _e.g._ **figure 1m**). Using all non-metastatic tumors as background, we found
very low levels of enrichment, less than $10\%$, in a handful of pathways. We
reasoned that the individual clusters might be too different to detect global
metastasis enrichment signal given the small sample size (n = 215 metastatic
tumor samples).

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 27 enriched pathways across all clusters (**table 2**). A number of
enrichments were found in multiple clusters represented by pathways that were
already shown to be enriched in a neighboring cluster. For example,
“Signaling by PTK6” is enriched in non-metastatic samples of cluster 8 (see 
**supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
metastatic tumors of clusters 7 and 9 (p < 10-3, **table 2**). This is also true of 
“Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
S4**) and also in metastatic tumors of clusters 7 and 9. <strike>Cluster 1, one of two clusters 
for which we could not identify any distinguishing pathways (cluster 8 being the other), 
had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other
clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).</strike> Cluster 4 metastatic
samples was enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathways, 
which are specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
specific to metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.

```{r child='../code/table2.Rmd'}
```
<mark>Statement does not apply anymore. The highest # of enriched path is 3. Not many are in 
metastasis. Only 2. “Fc epsilon receptor (FCERI) signaling” and "Signaling by SCF-KIT" </mark>
<strike>There were also a number of pathways that were enriched with lower specificity
(among 4 or more clusters) which were found to be enriched in metastases. We
identified 18 such pathways of which 12 had significantly higher mutation rates
in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
dependent signaling in response to WNT”, “Class I MHC mediated antigen
processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**). </strike>
